$Z^\prime$ mass limits and the naturalness of supersymmetry

### 2015-05-24 paper/Referee-response.tex

\documentclass[12pt]{article}
\usepackage{fullpage}
\begin{document}
\noindent Dear editor,
\\

\noindent Thanks for sending us the referee responses. We would like to thank
the referee for their helpful comments and questions regarding our
paper. In response we have made changes to our manuscript and feel
these have improved the paper, we therefore hope our paper will now be
accepted for publication in PRD.
\\
\noindent We respond to each point raised by the referee below, explaining
where appropriate what we have done to address it in the paper.

\begin{itemize}
\item {\it ``in eq. (9) an additional parameter $\mu'$ appears.  Where
does it come from?  Even though it is later fixed to 5 TeV, it is not
clear if it has any impact on their study?  Even if it has no impact,
this should be stated somewhere.''}

The referee is correct to point out that the bilinear coupling $\mu'$
is absent from the list of superpotential terms, and is lacking an
explanation in our original submission. The $\mu'$ term is associated
with the incomplete $27$ and $\overline{27}$ to which the fields
$\hat{H'}$ and $\hat{\overline{H}'}$ belong, but is invariant under
the Standard Model gauge group and the low energy $U(1)'$
symmetry.  Additionally, because this parameter does not appear in the
EWSB conditions the impact from it on the fine tuning can be
neglected, and so fixing the value of $\mu'$ does not impact the
results.

### 2015-05-24 paper/Referee-response.tex

\documentclass[12pt]{article}
\usepackage{fullpage}
\begin{document}
\noindent Dear editor,
\\

\noindent Thank you for sending us the referee report.  We would like to
sincerely thank the referee for their very helpful comments and questions
about our paper.  In response to each of these we have made appropriate
changes to our manuscript and feel these have improved the paper.  We
therefore hope that our paper will now be considered acceptable for
publication in Physical Review D.
\\
\noindent We respond to each point raised by the referee below, explaining
where appropriate what changes we have made in the paper to address the
referee's concerns.

\begin{itemize}
\item {\it ``in eq. (9) an additional parameter $\mu'$ appears.  Where
does it come from?  Even though it is later fixed to 5 TeV, it is not
clear if it has any impact on their study?  Even if it has no impact,
this should be stated somewhere.''}

The referee is correct to point out that the bilinear coupling $\mu'$
is absent from the list of superpotential terms, and is lacking an
explanation in our original submission. The $\mu'$ term is associated
with the incomplete $27'$ and $\overline{27}'$ to which the fields
$\hat{H'}$ and $\hat{\overline{H}'}$ belong, but is invariant under
the Standard Model gauge group and the low energy $U(1)'$
symmetry.  Additionally, because this parameter does not appear in the
EWSB conditions the impact from it on the fine tuning can be
neglected, and so fixing the value of $\mu'$ does not impact the
results.

### 2015-05-24 paper/e6ssm-tuning-paper.tex
Fixed some more minor typos

To quantify the fine tuning we will employ the traditional
Barbieri-Giudice measure \cite{Ellis:1986yg, Barbieri:1987fn}.  This
has been used extensively within the literature
e.g. \cite{deCarlos:1993yy, deCarlos:1993ca, Chankowski:1997zh,
  Agashe:1997kn, Wright:1998mk, Kane:1998im, BasteroGil:1999gu,
  Feng:1999zg, Allanach:2000ii, Dermisek:2005ar, Barbieri:2005kf,
  Allanach:2006jc, Gripaios:2006nn, Dermisek:2006py, Barbieri:2006dq,
  Kobayashi:2006fh, Perelstein:2012qg, Antusch:2012gv,Cheng:2012pe,
  CahillRowley:2012rv, Ross:2012nr, Basak:2012bd, Kang:2012sy,
  Athron:2013ipa, Boehm:2013qva, Miller:2013jra, Binjonaid:2014oga,
  Miller:2014jza}.

A number of alternative measures have also been applied in the
literature \cite{Anderson:1994dz, Anderson:1994tr, Anderson:1995cp,
Anderson:1996ew, Ciafaloni:1996zh, Chan:1997bi, Barbieri:1998uv,
Giusti:1998gz, Casas:2003jx, Casas:2004uu, Casas:2004gh,
Casas:2006bd, Kitano:2005wc, Athron:2007ry, Athron:2007qr,
Baer:2012up} with varying motivations.  A very different approach is
to work within a Bayesian analysis.  There the concept of naturalness
is automatically incorporated since in models where one must fine tune
parameters to fit measured values of the observables, the region with
high likelihood will occupy a tiny prior volume \cite{Allanach:2007qk,
Cabrera:2008tj, Ghilencea:2012gz, Ghilencea:2012qk, Fichet:2012sn,
Kim:2013uxa} suppressing the posterior.  Indeed in the MSSM and
the next-to-MSSM (NMSSM) if one transforms GUT scale parameters to the
VEVs, the inverse of the Jacobian for this transformation looks quite like
the derivatives that appear in the traditional fine tuning measure
\cite{Allanach:2007qk, Cabrera:2008tj, Kim:2013uxa}.  If one thinks
more generally then a model without fine tuning is one where the
parameterization is such that all the parameters are observables
\cite{Fichet:2012sn, Kim:2013uxa}.  This provides a quite general
definition of fine tuning as $1/|J|$ where $|J|$ is the determinant of
the Jacobian for the co-ordinate transformation between the parameters
and the observables.  Interestingly this means the tuning is the ratio
of the infinitesimal observable space space volume element to the
infinitesimal parameter space element and coincides with the measure
proposed in \cite{Athron:2007ry} when the interval of variation is
taken to zero.

@@

The breaking of $E_6$ into $SO(10)$, gives rise to $E_6 \to
SO(10) \times U(1)_{\psi}$, and the subsequent breaking of $SO(10)$
into $SU(5)$, gives $SO(10) \to SU(5) \times U(1)_{\chi}$ (this is
reviewed in e.g.~\cite{Langacker:2008yv}).  The extra $U(1)$ gauge
symmetry at low energies should then be a linear combination of these
in the $E_6$ inspired case, \be U(1)^\prime = U(1)_{\chi} \cos\theta
+ U(1)_{\psi} \sin\theta .  \ee In Table \ref{tab:E6charges} the
charges for several popular $E_6$ inspired $U(1)$ extensions are
shown.

@@

$U(1)$ and $E_6$ inspired extensions of the MSSM have been studied
very widely in the literature \cite{Gunion:1989we, Gunion:1992hs,
Binetruy:1985xm, Ellis:1986yg, Ibanez:1986si, Gunion:1986ky,
Haber:1986gz, Baer:1987eb, Gunion:1987jd, Grifols:1986vr,
Ellis:1986ip, Morris:1987fm, Drees:1987tp, Ma:1995xk,
Suematsu:1997tv, Suematsu:1997qt, Suematsu:1997au, Keith:1996fv,
Keith:1997zb, Gherghetta:1996yr, Demir:1998dk, Langacker:1998tc,
Hambye:2000bn, Ma:2000jf} (or for reviews see \cite{Hewett:1988xc,
Langacker:2008yv}).  There has also been a lot of work recently including
investigations of the neutralino sector \cite{Hesselbach:2001ri,
Barger:2005hb, Choi:2006fz, Barger:2007nv}; the relic density of dark matter
\cite{Kalinowski:2008iq}; GUT scale family symmetries which can explain the
hierarchy of masses in the fermion sector and their associated mixings
\cite{Stech:2008wd}; neutrino physics \cite{Kang:2004ix}; explanations of
the matter-anti-matter asymmetry of the universe though EW baryogenesis or
leptogenesis \cite{Hambye:2000bn, Ma:2000jf, Kang:2004pp}; decays of the
$Z^\prime$ boson \cite{Kang:2004bz, Baumgart:2006pa, Athron:2011wu,
Chang:2011be}; dipole moments \cite{GutierrezRodriguez:2006hb}; anomaly
mediated SUSY breaking with D-term contributions \cite{Asano:2008ju} and
the (extended) Higgs sectors \cite{Daikoku:2000ep, Ham:2008xf}.

@@

The impact of gauge kinetic mixing in the case where both of the extra
$U(1)$ symmetries appearing from the breakdown of $E_6$ are present at
low energy was studied in \cite{Rizzo:2012rf}.  The E$_6$SSM was also
included in studies looking at how first or second generation sfermion
masses can be used to constrain the GUT scale parameters \cite{Miller:2012vn}
and the renormalization of VEVs \cite{Sperling:2013eva, Sperling:2013xqa}.
The particle spectrum and collider signatures of the cE$_6$SSM have been
studied in a series of papers, \cite{Athron:2009ue, Athron:2009bs,
Athron:2011wu,Athron:2012sq}.  The threshold corrections to the
$\overline{DR}$ gauge and Yukawa couplings in the E$_6$SSM have also been
calculated and their numerical impact in the constrained version examined
\cite{Athron:2012pw}.

@@

As was noted in \cite{Athron:2013ipa}, the first two of the conditions
Eq.~\ref{eq:E6EWSBConditions} may be rewritten in the form
\begin{equation} \label{eq:E6MZequation}
\frac{M_Z^2}{2} = -\frac{\lambda^2 s^2}{2} + \frac{\tilde{m}_{H_d}^2 -
\tilde{m}_{H_u}^2 \tan^2\beta}{\tan^2\beta - 1} +
\frac{D_{H_d} - D_{H_u} \tan^2\beta}{\tan^2\beta - 1} ,
\end{equation}
\begin{equation} \label{eq:E6sin2bequation}
\sin 2\beta = \frac{\sqrt{2} \lambda A_{\lambda} s}
{\tilde{m}_{H_d}^2 + \tilde{m}_{H_u}^2 + \lambda^2 s^2 + D_{H_d} + D_{H_u}} ,
\end{equation}
with $M_Z^2 = \bar{g}^2 v^2 / 4$, $v^2 = v_1^2 + v_2^2$ and $\tan\beta =
v_2 / v_1$ and where we have for convenience absorbed the effects of the
loop corrections into the soft masses,
\begin{align*}
\tilde{m}_{H_d}^2 &= m_{H_d}^2 + \frac{1}{v_1} \frac{\partial \Delta V}
{\partial v_1} , \\
\tilde{m}_{H_u}^2 &= m_{H_u}^2 + \frac{1}{v_2} \frac{\partial \Delta V}
{\partial v_2} .
\end{align*}
Written in the form of Eq.~(\ref{eq:E6MZequation}) we see the
potential new source of fine tuning alluded to above, in the form of
the third term on the right-hand side.  For large values of the VEV
$s$, the $D$-term contributions can be quite a bit larger than
$M_Z^2$.  In particular, recent experimental limits \cite{Aad:2014cka}
require that the $Z'$ mass be large, with for example bounds of
$M_{Z'} \gtrsim 2.51$ TeV in $U(1)_\psi$ models and $M_{Z'} \gtrsim 2.62$ TeV
in $U(1)_\chi$ models.  To satisfy these limits typically requires large
values of the singlet VEV $s$.  For example $s \gtrsim 6$ TeV is required in
the E$_6$SSM with $U(1)' = U(1)_N$, so that $|D_{H_1}| , |D_{H_2}| \gg M_Z^2$
for $E_6$ models with $Q_S \neq 0$.  As a result the remaining terms on the
right-hand side of Eq.~(\ref{eq:E6MZequation}) must be tuned to cancel
this very large contribution to $M_Z$.  Moreover, because this is a large tree
level fine tuning, it may negate the improvement in naturalness that is
associated with having a reduced need for heavy superpartners.  In $U(1)$
extended models for which $Q_S\neq 0$, the importance of the $Z^\prime$ mass
to the fine tuning in these models can be made even clearer by writing
Eq.~(\ref{eq:E6MZequation}) in the form given in \cite{Athron:2013ipa},
\begin{equation}\label{eq:originalE6MZequation}
c(\theta , \tan\beta) \frac{M_Z^2}{2} = -\frac{\lambda^2 s^2}{2} +
\frac{\tilde{m}_{H_d}^2 - \tilde{m}_{H_u}^2 \tan^2\beta}{\tan^2\beta - 1}
+ d(\theta , \tan\beta) \frac{M_{Z^\prime}^2}{2} ,
\end{equation}
where
\begin{align}
c(\theta , \tan\beta) &= 1 - \frac{4}{\left (\tan^2\beta - 1 \right )}
\frac{g_1'^2}{\bar{g}^2} \left (Q_1 - Q_2 \tan^2\beta \right )
\left (Q_1 \cos^2\beta + Q_2 \sin^2\beta \right ) , \label{eq:cdefn} \\
d(\theta , \tan\beta) &= \frac{Q_1 - Q_2 \tan^2\beta}
{Q_S \left (\tan^2\beta - 1\right )} . \label{eq:ddefn}
\end{align}
Written like so it is evident that the fine tuning contribution coming
from the new $D$-terms depends both on the $U(1)'$ charges and the
$Z^\prime$ mass, and that the tuning can be expected to increase with
$M_{Z^\prime}$.  As shall be shown below, the exact size of this tuning
then depends strongly on the choice of $U(1)'$ charges, via the
coefficient $d$.

@@

However if the extra $U(1)$ gauge symmetries appears from the
breakdown of $E_6$, then $\sin\chi$ should be zero at the GUT scale.
Nonetheless even if this term is zero at the outset it will still be
radiatively generated if the trace of their charges, $\sum_i Q_i^Y
Q^\prime_i$, is non-zero.  In the cases studied here the trace of the
charges over states in the complete $27$-plets vanish, but to be
consistent with single step gauge coupling unification we also
included $\hat{H}^\prime$ and $\hat{\overline{H^\prime}}$ which lead
to a non-zero value for $\sum_i Q_i^Y Q^\prime_i $.  The value induced
by this at the EW scale though is rather small as can be
seen\footnote{The specific incomplete multiplets we consider here
  correspond to the third of the four possible embeddings referred to in
  Ref.~\cite{Rizzo:1998ut}.} in Fig.~3 of Ref.~\cite{Rizzo:1998ut}, and this
was also checked
with two-loop RGEs in the E$_6$SSM \cite{King:2005jy,Athron:2009bs}.
For this reason and due to the huge expansion in the number of terms
in the two-loop RGEs when one allows for gauge kinetic mixing we will
neglect this in our analysis here and throughout this paper.

In general though it is possible for gauge kinetic mixing to be much larger
as can be the case if one considers an additional $5 + \overline{5}$ pair of
$SU(5)$ multiplets \cite{Rizzo:1998ut} or as has been looked in the $U(1)_{B-L}$
\cite{Salvioni:2009mt,Krauss:2012ku}.  In such a case this will impact the
results in two ways, firstly by altering the $Z^\prime$ limit from experiment
and secondly by altering the charges which appear in the EWSB condition,
which can be seen from examining
Eqs.~(\ref{eq:originalE6MZequation})-(\ref{eq:ddefn}).

@@

As stated above, to quantify the resulting fine tuning we apply the
traditional Barbieri-Giudice measure \cite{Ellis:1986yg, Barbieri:1987fn}.
A specific model is characterized by a set of $n$ model parameters $\{p_i\}$
and is defined at some input scale $M_X$. For a given parameter $p$ in this
set, one computes an associated sensitivity
\begin{equation} \label{eq:bgmeasure}
\Delta_p = \left | \frac{\partial \ln M_Z^2}{\partial \ln p} \right
| = \left | \frac{p}{M_Z^2} \frac{\partial M_Z^2}{\partial p} \right | .
\end{equation}
The coefficient $\Delta_p$ measures the fractional variation in $M_Z^2$
resulting from a given variation in the parameter $p$.  The overall fine
tuning is then taken to be
$\Delta = \underset{i}{\max} \, \{ \Delta_{p_i} \}$.\\

The sensitivity $\Delta_p$ may be calculated directly from the expression
for $M_Z^2$ in terms of the $p_i$ for a particular model, which leads to a
so-called master formula for calculating the fine tuning.  A master formula
for the E$_6$SSM, obtained from the tree level scalar potential, was
presented in \cite{Athron:2013ipa}.  In order to derive the expression
presented there, the fact that $s \gg v$ was made use of to neglect certain
$O(v^2)$ terms in the EWSB conditions, greatly simplifying the final result.
For the purposes of exploring a wider class of $E_6$ inspired models, we have
derived the master formula without neglecting any $O(v^2)$ terms.  The more
complete tree-level master formula is somewhat complicated.  This is
because, unlike in the MSSM, even at tree level it is not possible to
solve explicitly for the VEVs $v_1$, $v_2$ in terms of the Lagrangian
parameters.  It may be written in the form \be \Delta_p = |C|^{-1} \times
\frac{|p|}{M_Z} \bigg | \sum_{q} \tilde{\Delta}_q \frac{\partial q}
{\partial p} \bigg | , \label{eq:E6masterformula}
\ee where the sum is over all low energy running parameters appearing
in the tree level EWSB conditions, i.e., $q \in \{\lambda , A_\lambda ,
m_{H_d}^2 , m_{H_u}^2 , m_S^2 , g_1 , g_2 , g_1'\}$.  Expressions for the
quantity $C$ and the $\tilde{\Delta}_q$ appearing above are given in
Appendix \ref{app:masterformula}.  It should be noted that the effects
of $U(1)$ mixing are neglected in deriving Eq.~(\ref{eq:E6masterformula}).

>>>>

To quantify the fine tuning we will employ the traditional
Barbieri-Giudice measure \cite{Ellis:1986yg, Barbieri:1987fn}.  This
has been used extensively within the literature
e.g. Refs.~\cite{deCarlos:1993yy, deCarlos:1993ca, Chankowski:1997zh,
  Agashe:1997kn, Wright:1998mk, Kane:1998im, BasteroGil:1999gu,
  Feng:1999zg, Allanach:2000ii, Dermisek:2005ar, Barbieri:2005kf,
  Allanach:2006jc, Gripaios:2006nn, Dermisek:2006py, Barbieri:2006dq,
  Kobayashi:2006fh, Perelstein:2012qg, Antusch:2012gv,Cheng:2012pe,
  CahillRowley:2012rv, Ross:2012nr, Basak:2012bd, Kang:2012sy,
  Athron:2013ipa, Boehm:2013qva, Miller:2013jra, Binjonaid:2014oga,
  Miller:2014jza}.

A number of alternative measures have also been applied in the
literature \cite{Anderson:1994dz, Anderson:1994tr, Anderson:1995cp,
Anderson:1996ew, Ciafaloni:1996zh, Chan:1997bi, Barbieri:1998uv,
Giusti:1998gz, Casas:2003jx, Casas:2004uu, Casas:2004gh,
Casas:2006bd, Kitano:2005wc, Athron:2007ry, Athron:2007qr,
Baer:2012up} with varying motivations.  A very different approach is
to work within a Bayesian analysis.  There the concept of naturalness
is automatically incorporated since in models where one must fine tune
parameters to fit measured values of the observables, the region with
high likelihood will occupy a tiny prior volume \cite{Allanach:2007qk,
Cabrera:2008tj, Ghilencea:2012gz, Ghilencea:2012qk, Fichet:2012sn,
Kim:2013uxa} suppressing the posterior.  Indeed in the MSSM and
the next-to-MSSM (NMSSM) if one transforms GUT scale parameters to the
VEVs, the inverse of the Jacobian for this transformation looks quite like
the derivatives that appear in the traditional fine tuning measure
\cite{Allanach:2007qk, Cabrera:2008tj, Kim:2013uxa}.  If one thinks
more generally then a model without fine tuning is one where the
parameterization is such that all the parameters are observables
\cite{Fichet:2012sn, Kim:2013uxa}.  This provides a quite general
definition of fine tuning as $1/|J|$ where $|J|$ is the determinant of
the Jacobian for the co-ordinate transformation between the parameters
and the observables.  Interestingly this means the tuning is the ratio
of the infinitesimal observable space volume element to the
infinitesimal parameter space element and coincides with the measure
proposed in Ref.~\cite{Athron:2007ry} when the interval of variation is
taken to zero.

@@

The breaking of $E_6$ into $SO(10)$, gives rise to $E_6 \to
SO(10) \times U(1)_{\psi}$, and the subsequent breaking of $SO(10)$
into $SU(5)$, gives $SO(10) \to SU(5) \times U(1)_{\chi}$ (this is
reviewed in e.g. Ref.~\cite{Langacker:2008yv}).  The extra $U(1)$ gauge
symmetry at low energies should then be a linear combination of these
in the $E_6$ inspired case, \be U(1)^\prime = U(1)_{\chi} \cos\theta
+ U(1)_{\psi} \sin\theta .  \ee In Table \ref{tab:E6charges} the
charges for several popular $E_6$ inspired $U(1)$ extensions are
shown.

@@

$U(1)$ and $E_6$ inspired extensions of the MSSM have been studied
very widely in the literature \cite{Gunion:1989we, Gunion:1992hs,
Binetruy:1985xm, Ellis:1986yg, Ibanez:1986si, Gunion:1986ky,
Haber:1986gz, Baer:1987eb, Gunion:1987jd, Grifols:1986vr,
Ellis:1986ip, Morris:1987fm, Drees:1987tp, Ma:1995xk,
Suematsu:1997tv, Suematsu:1997qt, Suematsu:1997au, Keith:1996fv,
Keith:1997zb, Gherghetta:1996yr, Demir:1998dk, Langacker:1998tc,
Hambye:2000bn, Ma:2000jf} (or for reviews see Refs.~\cite{Hewett:1988xc,
Langacker:2008yv}).  There has also been a lot of work recently including
investigations of the neutralino sector \cite{Hesselbach:2001ri,
Barger:2005hb, Choi:2006fz, Barger:2007nv}; the relic density of dark matter
\cite{Kalinowski:2008iq}; GUT scale family symmetries which can explain the
hierarchy of masses in the fermion sector and their associated mixings
\cite{Stech:2008wd}; neutrino physics \cite{Kang:2004ix}; explanations of
the matter-anti-matter asymmetry of the universe though EW baryogenesis or
leptogenesis \cite{Hambye:2000bn, Ma:2000jf, Kang:2004pp}; decays of the
$Z^\prime$ boson \cite{Kang:2004bz, Baumgart:2006pa, Athron:2011wu,
Chang:2011be}; dipole moments \cite{GutierrezRodriguez:2006hb}; anomaly
mediated SUSY breaking with D-term contributions \cite{Asano:2008ju} and
the (extended) Higgs sectors \cite{Daikoku:2000ep, Ham:2008xf}.

@@

The impact of gauge kinetic mixing in the case where both of the extra
$U(1)$ symmetries appearing from the breakdown of $E_6$ are present at
low energy was studied in Ref.~\cite{Rizzo:2012rf}.  The E$_6$SSM was also
included in studies looking at how first or second generation sfermion
masses can be used to constrain the GUT scale parameters \cite{Miller:2012vn}
and the renormalization of VEVs \cite{Sperling:2013eva, Sperling:2013xqa}.
The particle spectrum and collider signatures of the cE$_6$SSM have been
studied in a series of papers, \cite{Athron:2009ue, Athron:2009bs,
Athron:2011wu,Athron:2012sq}.  The threshold corrections to the
$\overline{DR}$ gauge and Yukawa couplings in the E$_6$SSM have also been
calculated and their numerical impact in the constrained version examined
\cite{Athron:2012pw}.

@@

As was noted in Ref.~\cite{Athron:2013ipa}, the first two of the conditions in
Eq.~(\ref{eq:E6EWSBConditions}) may be rewritten in the form
\begin{equation} \label{eq:E6MZequation}
\frac{M_Z^2}{2} = -\frac{\lambda^2 s^2}{2} + \frac{\tilde{m}_{H_d}^2 -
\tilde{m}_{H_u}^2 \tan^2\beta}{\tan^2\beta - 1} +
\frac{D_{H_d} - D_{H_u} \tan^2\beta}{\tan^2\beta - 1} ,
\end{equation}
\begin{equation} \label{eq:E6sin2bequation}
\sin 2\beta = \frac{\sqrt{2} \lambda A_{\lambda} s}
{\tilde{m}_{H_d}^2 + \tilde{m}_{H_u}^2 + \lambda^2 s^2 + D_{H_d} + D_{H_u}} ,
\end{equation}
with $M_Z^2 = \bar{g}^2 v^2 / 4$, $v^2 = v_1^2 + v_2^2$ and $\tan\beta =
v_2 / v_1$ and where we have for convenience absorbed the effects of the
loop corrections into the soft masses,
\begin{align*}
\tilde{m}_{H_d}^2 &= m_{H_d}^2 + \frac{1}{v_1} \frac{\partial \Delta V}
{\partial v_1} , \\
\tilde{m}_{H_u}^2 &= m_{H_u}^2 + \frac{1}{v_2} \frac{\partial \Delta V}
{\partial v_2} .
\end{align*}
Written in the form of Eq.~(\ref{eq:E6MZequation}) we see the
potential new source of fine tuning alluded to above, in the form of
the third term on the right-hand side.  For large values of the VEV
$s$, the $D$-term contributions can be quite a bit larger than
$M_Z^2$.  In particular, recent experimental limits \cite{Aad:2014cka}
require that the $Z'$ mass be large, with for example bounds of
$M_{Z'} \gtrsim 2.51$ TeV in $U(1)_\psi$ models and $M_{Z'} \gtrsim 2.62$ TeV
in $U(1)_\chi$ models.  To satisfy these limits typically requires large
values of the singlet VEV $s$.  For example $s \gtrsim 6$ TeV is required in
the E$_6$SSM with $U(1)' = U(1)_N$, so that $|D_{H_1}| , |D_{H_2}| \gg M_Z^2$
for $E_6$ models with $Q_S \neq 0$.  As a result the remaining terms on the
right-hand side of Eq.~(\ref{eq:E6MZequation}) must be tuned to cancel
this very large contribution to $M_Z$.  Moreover, because this is a large tree
level fine tuning, it may negate the improvement in naturalness that is
associated with having a reduced need for heavy superpartners.  In $U(1)$
extended models for which $Q_S\neq 0$, the importance of the $Z^\prime$ mass
to the fine tuning in these models can be made even clearer by writing
Eq.~(\ref{eq:E6MZequation}) in the form given in Ref.~\cite{Athron:2013ipa},
\begin{equation}\label{eq:originalE6MZequation}
c(\theta , \tan\beta) \frac{M_Z^2}{2} = -\frac{\lambda^2 s^2}{2} +
\frac{\tilde{m}_{H_d}^2 - \tilde{m}_{H_u}^2 \tan^2\beta}{\tan^2\beta - 1}
+ d(\theta , \tan\beta) \frac{M_{Z^\prime}^2}{2} ,
\end{equation}
where
\begin{align}
c(\theta , \tan\beta) &= 1 - \frac{4}{\left (\tan^2\beta - 1 \right )}
\frac{g_1'^2}{\bar{g}^2} \left (Q_1 - Q_2 \tan^2\beta \right )
\left (Q_1 \cos^2\beta + Q_2 \sin^2\beta \right ) , \label{eq:cdefn} \\
d(\theta , \tan\beta) &= \frac{Q_1 - Q_2 \tan^2\beta}
{Q_S \left (\tan^2\beta - 1\right )} . \label{eq:ddefn}
\end{align}
Written like so it is evident that the fine tuning contribution coming
from the new $D$-terms depends both on the $U(1)'$ charges and the
$Z^\prime$ mass, and that the tuning can be expected to increase with
$M_{Z^\prime}$.  As shall be shown below, the exact size of this tuning
then depends strongly on the choice of $U(1)'$ charges, via the
coefficient $d$.

@@

However if the extra $U(1)$ gauge symmetry appears from the
breakdown of $E_6$, then $\sin\chi$ should be zero at the GUT scale.
Nonetheless even if this term is zero at the outset it will still be
radiatively generated if the trace of the $U(1)$ charges, $\sum_i Q_i^Y
Q^\prime_i$, is non-zero.  In the cases studied here the trace of the
charges over states in the complete $27$-plets vanish, but to be
consistent with single step gauge coupling unification we also
included $\hat{H}^\prime$ and $\hat{\overline{H^\prime}}$ which lead
to a non-zero value for $\sum_i Q_i^Y Q^\prime_i $.  The value induced
by this at the EW scale though is rather small as can be
seen\footnote{The specific incomplete multiplets we consider here
  correspond to the third of the four possible embeddings referred to in
  Ref.~\cite{Rizzo:1998ut}.} in Fig.~3 of Ref.~\cite{Rizzo:1998ut}, and this
was also checked
with two-loop RGEs in the E$_6$SSM \cite{King:2005jy,Athron:2009bs}.
For this reason and due to the huge expansion in the number of terms
in the two-loop RGEs when one allows for gauge kinetic mixing we will
neglect this in our analysis here and throughout this paper.

In general though it is possible for gauge kinetic mixing to be much larger
as can be the case if one considers an additional $5 + \overline{5}$ pair of
$SU(5)$ multiplets \cite{Rizzo:1998ut} or as has been looked at in the
$U(1)_{B-L}$ \cite{Salvioni:2009mt,Krauss:2012ku}.  In such a case this will
impact the results in two ways, firstly by altering the $Z^\prime$ limit from
experiment and secondly by altering the charges which appear in the EWSB
condition, which can be seen from examining
Eqs.~(\ref{eq:originalE6MZequation})-(\ref{eq:ddefn}).

@@

As stated above, to quantify the resulting fine tuning we apply the
traditional Barbieri-Giudice measure \cite{Ellis:1986yg, Barbieri:1987fn}.
A specific model is characterized by a set of $n$ model parameters $\{p_i\}$
and is defined at some input scale $M_X$. For a given parameter $p$ in this
set, one computes an associated sensitivity
\begin{equation} \label{eq:bgmeasure}
\Delta_p = \left | \frac{\partial \ln M_Z^2}{\partial \ln p} \right
| = \left | \frac{p}{M_Z^2} \frac{\partial M_Z^2}{\partial p} \right | .
\end{equation}
The coefficient $\Delta_p$ measures the fractional variation in $M_Z^2$
resulting from a given variation in the parameter $p$.  The overall fine
tuning is then taken to be
$\Delta = \underset{i}{\max} \, \{ \Delta_{p_i} \}$.

The sensitivity $\Delta_p$ may be calculated directly from the expression
for $M_Z^2$ in terms of the $p_i$ for a particular model, which leads to a
so-called master formula for calculating the fine tuning.  A master formula
for the E$_6$SSM, obtained from the tree level scalar potential, was
presented in Ref.~\cite{Athron:2013ipa}.  In order to derive the expression
presented there, the fact that $s \gg v$ was made use of to neglect certain
$O(v^2)$ terms in the EWSB conditions, greatly simplifying the final result.
For the purposes of exploring a wider class of $E_6$ inspired models, we have
derived the master formula without neglecting any $O(v^2)$ terms.  The more
complete tree-level master formula is somewhat complicated.  This is
because, unlike in the MSSM, even at tree level it is not possible to
solve explicitly for the VEVs $v_1$, $v_2$ in terms of the Lagrangian
parameters.  It may be written in the form \be \Delta_p = |C|^{-1} \times
\frac{|p|}{M_Z} \bigg | \sum_{q} \tilde{\Delta}_q \frac{\partial q}
{\partial p} \bigg | , \label{eq:E6masterformula}
\ee where the sum is over all low energy running parameters appearing
in the tree level EWSB conditions, i.e., $q \in \{\lambda , A_\lambda ,
m_{H_d}^2 , m_{H_u}^2 , m_S^2 , g_1 , g_2 , g_1'\}$.  Expressions for the
quantity $C$ and the $\tilde{\Delta}_q$ appearing above are given in
Appendix \ref{app:masterformula}.  It should be noted that the effects
of $U(1)$ mixing are neglected in deriving Eq.~(\ref{eq:E6masterformula}).

However, it is well known in the MSSM that radiative corrections can
significantly change (indeed, reduce) the fine tuning \cite{Cassel:2010px}.
It is therefore important when studying the fine tuning to include loop
corrections to the effective potential in the fine tuning measure.  To do so
it is most convenient to work in terms of the EWSB conditions
Eq.~(\ref{eq:E6EWSBConditions}), rather than Eq.~(\ref{eq:E6MZequation}).
The general procedure that we use is as follows (this method has also
previously been applied in the NMSSM, see for example
Ref.~\cite{Ellwanger:2011mu}).  For a model in which $m$ fields develop real
VEVs (e.g. $m=2$ in the MSSM, $m=3$ in the NMSSM and in the $E_6$
models considered), we require that the $m$ minimization conditions
\begin{equation} \label{eq:EWSBconditions}
f_1 = f_2 = \dotsb = f_m = 0
\end{equation}
continue to hold under an arbitrary variation in a model parameter
$p \rightarrow p + \delta p$, so that the variations $\delta f_i$ satisfy
\begin{equation} \label{eq:EWSBvariations}
\delta f_1 = \delta f_2 = \dotsb = \delta f_m = 0 .
\end{equation}
Each $f_i$ is a function of the VEVs $v_j$ and $l$ running parameters
$q_k$ evaluated at the scale of EWSB, $f_i = f_i(v_j , q_k)$.  Thus for
each $f_i$ we have that
\begin{equation} \label{eq:EWSBchainrule}
\delta f_i = \sum_{j=1}^m \frac{\partial f_i}{\partial v_j}
\frac{\partial v_j}{\partial p} + \sum_{k=1}^l \frac{\partial f_i}
{\partial q_k} \frac{\partial q_k}{\partial p} = 0 .
\end{equation}
The quantities $\frac{\partial f_i}{\partial v_j}$ are the elements of
the CP-even Higgs squared mass matrix $M_h^2$ of the model before
rotating into the mass eigenstate basis.  When evaluated for all $n$
model parameters, the above system of equations can be concisely
expressed as
\begin{equation} \label{eq:tuningsystem}
M_h^2 \begin{pmatrix}
\frac{\partial v_1}{\partial p_1} & \cdots &
\frac{\partial v_1}{\partial p_n} \\
\vdots & \ddots & \vdots \\
\frac{\partial v_m}{\partial p_1} & \cdots & \frac{\partial v_m}{\partial p_n}
\end{pmatrix} = -\begin{pmatrix}
\frac{\partial f_1}{\partial q_1} & \cdots & \frac{\partial f_1}
{\partial q_l} \\
\vdots & \ddots & \vdots \\
\frac{\partial f_m}{\partial q_1} & \cdots & \frac{\partial f_m}{\partial q_l}
\end{pmatrix}
\begin{pmatrix}
\frac{\partial q_1}{\partial p_1} & \cdots & \frac{\partial q_1}
{\partial p_n} \\
\vdots & \ddots & \vdots \\
\frac{\partial q_l}{\partial p_1} & \cdots & \frac{\partial q_l}{\partial p_n}
\end{pmatrix} .
\end{equation}
The quantities forming the first matrix on the right-hand side, along
with $M_h^2$, are easily calculated by differentiating the conditions
in Eq.~(\ref{eq:E6EWSBConditions}) with respect to the VEVs and the
running parameters.  The remaining derivatives $\partial q_k /\partial
p$ must be determined using the RGEs.  Once these have been obtained,
it is straightforward to solve for the $\partial v_i / \partial p$.
The sensitivities $\Delta_p$ are then simply linear combinations of
the $\partial v_i / \partial p$ and $\partial q_k / \partial p$.  The
effects of radiative corrections may be easily included by including
the Coleman-Weinberg potential contributions $\Delta V$ in the EWSB
conditions.  Here we use the one-loop corrections given in
Eq.~(\ref{Eq:DeltaV}).
